---
title: "Deseq analysis comparing between days of gene expression"
author: "Yubin Xie"
format: 
  html:
    self-contained: true
editor: visual
echo: True
---

```{r setup}
#| output: False
# import packages
library(readxl)
library(tidyverse)
library(writexl) 
library(ggthemes)
library(DESeq2)
library(reshape2)
library(edgeR)
```

Loading in the Data

```{r}
gene_expression <- read_excel("../B_Cell_Gene_Expression_Data.xlsx",  col_types = c("text", "text", "skip", "skip", rep("numeric", 18)))
gene_expression <- column_to_rownames(gene_expression, "GeneSymbol")
```

## Creating a Simple Visual Plot comparing day0 and day1 genetic concentration

```{r eval=FALSE}
#| title: "Simple Plot"
#| fig-cap: "We see a more of a dip in the B-cell gene expression level after 24 hours compare to none b-cell genes"

ggplot(gene_expression, aes(x = s1_Apheresis_day0, y = s1_Apheresis_day1)) +
  geom_point(aes(color="B-cell?")) + geom_smooth(method='lm') +
  labs(title = "Gene Expression", subtitle = "More words", x="Day0 for patient one", y= "Day1 for patient one")
  scale_color_colorblind()
  
```

```{r}
# Removes the b_cells column to keep only gene expression
gene_expression <- gene_expression[, -1]
```

## Checking if Data is Already Log2p Transformed

```{r}
hist(as.numeric(gene_expression[1,]))
qqnorm(as.numeric(gene_expression[1,]))
qqline(as.numeric(gene_expression[1, ]))
```

Not log transformed at all, looking at the graph. We see huge tails...

## Metadata dataframe for gene_expression

```{r}
# Create the metadata dataframe
colData <- data.frame(
  row.names = colnames(gene_expression),
  Sample = factor(rep(c(1,2,3),times=6)),
  Day = factor(rep(c("Day0","Day1","Day2"), each = 3, times = 2)),
  Condition = factor(rep(c("Apheresis", "Control"), each = 9))
  
)

# A column for two factor combinations 
colData$Condition_Day <- factor(paste0(colData$Condition, "_", colData$Day))
colData$Sample_Day <- factor(paste0(colData$Sample, "_", colData$Day))
colData$Sample_Condition <-  factor(paste0(colData$Sample, "_", colData$Condition))

# Explicitly set Control and Day0 as the baseline references
colData$Condition <- relevel(colData$Condition, ref = "Control")
colData$Day <- relevel(colData$Day, ref = "Day0")
colData$Condition_Day <- relevel(colData$Condition_Day, ref = "Control_Day0")

# Combining everything into one dataframe
x <- DGEList(gene_expression, samples = colData)
```

Since we don't have raw count, we can not use voom. We will try to log transform it ourself.

```{r}
exprs_data <- as.matrix(x$counts)
if(max(exprs_data, na.rm=TRUE) > 50) {
  exprs_data <- log2(exprs_data + 1)
}

```

## Visualizing the current clustering

```{r}
library(limma)
library(Glimma) 
# unsupervised Clustering plot to see if the grouping on PCA space make sense. RUN THIS TO GET A FEEL OF THE DATA AND SEE IF ANY BATCH CORRECTIONS ARE NESSESARY
glMDSPlot(exprs_data, group = x$samples, launch = TRUE)

```

## OK we definitely need batch correction since the samples from the same individual are domination and clustering at very similar PC coordinates

Solution: Include Sample as the batching factor

### Creating Design matrix for day and condition.

```{r}

# Creating factor arrays
Condition_Day <- x$samples$Condition_Day
Sample <- x$samples$Sample
Day <- x$samples$Day
design <- model.matrix(~Sample + Condition_Day) # Added Sample as batching factor

colnames(design) <- c(levels(Condition_Day), paste0("Sample", levels(Sample)[-1]))
colnames(design) <- gsub("Condition_Day", "", colnames(design))
colnames(design)[1] <- "Control_Day0" 
```

### Contrast matrix

```{r}
contr.matrix <- makeContrasts(
  Apheresis_Day2vsDay1 = Apheresis_Day2 - Apheresis_Day1,
  Apheresis_Day1vsDay0 = Apheresis_Day1 - Apheresis_Day0,
  Apheresis_Day2vsDay0 = Apheresis_Day2 - Apheresis_Day0,
  Control_Day2vsDay1 = Control_Day2 - Control_Day1,
  Control_Day1vsDay0 = Control_Day1,
  Control_Day2vsDay0 = Control_Day2,
  Apheresis_Day0vsControl_Day0 = Apheresis_Day0,
  Apheresis_Day1vsControl_Day1 = Apheresis_Day1 - Control_Day1,
  Apheresis_Day2vsControl_Day2 = Apheresis_Day2 - Control_Day2,
  levels=colnames(design)
)
```

```{r}
vfit <- lmFit(exprs_data, design)
vfit <- contrasts.fit(vfit, contrast = contr.matrix) 
efit <- eBayes(vfit, trend = TRUE) 
plotSA(efit)
```

```{r}

top_genes <- topTable(efit, coef = "Apheresis_Day0vsControl_Day0", adjust.method = "BH", number = Inf)

top_genes$Gene <- rownames(top_genes)

# 2. Categorize genes into Up, Down, or Not Significant
top_genes$Expression <- "Not Significant"
top_genes$Expression[top_genes$logFC > 1 & top_genes$adj.P.Val < 0.05] <- "Up-regulated"
top_genes$Expression[top_genes$logFC < -1 & top_genes$adj.P.Val < 0.05] <- "Down-regulated"
```

```{r}
library(ggplot2)
library(ggrepel) # Keeps text labels from overlapping

ggplot(top_genes, aes(x = logFC, y = -log10(P.Value))) +
  # Draw the gene dots, color-coded by their category
  geom_point(aes(color = Expression), alpha = 0.8, size = 1.2) +
  
  # Custom color palette matching your categories
  scale_color_manual(values = c("Down-regulated" = "dodgerblue3", 
                                "Not Significant" = "gray70", 
                                "Up-regulated" = "firebrick3")) +
  
  # Add threshold line guides
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black", alpha = 0.5) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black", alpha = 0.5) +
  
  # Label the top 10 up-regulated genes automatically without overlapping
  geom_text_repel(
    data = head(top_genes[top_genes$Expression == "Up-regulated" | top_genes$Expression == "Down-regulated", ], 20),
    aes(label = Gene),
    size = 3,
    max.overlaps = 15
  ) +
  
  # Clean, minimal presentation theme
  theme_minimal() +
  labs(
    title = "Differential Gene Expression",
    subtitle = "Apheresis Day 0 vs Control Day 0",
    x = "Log2 Fold Change",
    y = "-Log10 Raw p-value",
    color = "Status"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "right"
  )

```

```{r}

top_genes <- topTable(efit, coef = "Apheresis_Day0vsControl_Day0", adjust.method = "BH", number = Inf)
significant_top_genes <- top_genes[top_genes$adj.P.Val < 0.05 & top_genes$logFC > 1,]
print(significant_top_genes)
#write_xlsx(cbind(Index = rownames(significant_top_genes), significant_top_genes), "~/Documents/School/Bio Stuff/Bios_20151/Bio_20251_Final_Project/Output_plot/Apheresis_Day0vsControl_Day0_Deg_Up.xlsx")

significant_down_genes <- top_genes[top_genes$adj.P.Val < 0.05 & top_genes$logFC < -1,]
print(significant_down_genes)
#write_xlsx(cbind(Index = rownames(significant_down_genes), significant_down_genes), "~/Documents/School/Bio Stuff/Bios_20151/Bio_20251_Final_Project/Output_plot/Apheresis_Day0vsControl_Day0_Deg_Down.xlsx")
```

## GO term analysis for Down Regulated Genes

```{r}
library(clusterProfiler)
library(org.Hs.eg.db) # For Human genome mapping

# Put the genes from your image into a character vector
down_genes <- c(
  "SPATA20", "FCRL5", "CLEC20A", "GSDMB", "CORO2B", "FAM177B", 
  "IGHV6.1", "RGPD1", "CNTNAP2", "DCLK2", "ZNF83", "IGHV3.66", 
  "DIPK1B", "NPIPA5", "CXorf65", "ZNF107", "SYN3", "PLGLB1", 
  "IGHV2.70", "SDK2", "DACT1", "FAM167A", "PLD6", "CRIM1", 
  "SMIM11A", "IGHV4.4"
)
down_genes <- gsub("\\.", "-", down_genes)


# Run the enrichment for Biological Processes (BP)
go_results <- enrichGO(
  gene          = down_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",   
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)

# View the mapped GO categories
print(head(go_results))

# Generate a quick barplot of the most enriched functions
barplot(go_results, showCategory = 10, title = "Down on Day0 GO Term Analysis")


```

```{r}
up_genes <- gsub("\\.", "-", rownames(significant_top_genes))


# Run the enrichment for Biological Processes (BP)
go_results <- enrichGO(
  gene          = up_genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "SYMBOL",
  ont           = "BP",   
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.05
)

# View the mapped GO categories
print(head(go_results))

# Generate a quick barplot of the most enriched functions
barplot(go_results, showCategory = 10, title = "Up on Day0 GO Term Analysis")
```
