---
title: "Deseq analysis comparing between days of gene expression"
author: "Yubin Xie"
format: 
  html:
    self-contained: true
editor: visual
echo: False
---

```{r setup}
#| output: False
# import packages
library(readxl)
library(tidyverse)
library(writexl) 
library(ggthemes)
library(DESeq2)
library(reshape2)
library(edgeR)
```

Loading in the Data

```{r}
gene_expression <- read_excel("../B_Cell_Gene_Expression_Data.xlsx",  col_types = c("text", "skip", "skip", "skip", rep("numeric", 18)))
gene_expression <- column_to_rownames(gene_expression, "GeneSymbol")
```

## Creating a Simple Visual Plot comparing day0 and day1 genetic concentration

```{r eval=FALSE}
#| title: "Simple Plot"
#| fig-cap: "We see a more of a dip in the B-cell gene expression level after 24 hours compare to none b-cell genes"

ggplot(gene_expression, aes(x = s1_Apheresis_day0, y = s1_Apheresis_day1)) +
  geom_point(aes(color=Is_B_Cell)) + geom_smooth(method='lm') +
  labs(title = "Gene Expression", subtitle = "More words", x="Day0 for patient one", y= "Day1 for patient one")
  scale_color_colorblind()
  
```

## Checking if Data is Already Log2p Transformed

```{r}
hist(as.numeric(gene_expression[1,]))
qqnorm(as.numeric(gene_expression[1,]))
qqline(as.numeric(gene_expression[1, ]))
```

## Metadata dataframe for gene_expression

```{r}
# Create the metadata dataframe
colData <- data.frame(
  row.names = colnames(gene_expression),
  Sample = factor(rep(c(1,2,3),times=6)),
  Day = factor(rep(c("Day0","Day1","Day2"), each = 3, times = 2)),
  Condition = factor(rep(c("Apheresis", "Control"), each = 9))
  
)

# A column for two factor combinations 
colData$Condition_Day <- factor(paste0(colData$Condition, "_", colData$Day))
colData$Sample_Day <- factor(paste0(colData$Sample, "_", colData$Day))
colData$Sample_Condition <-  factor(paste0(colData$Sample, "_", colData$Condition))

# Explicitly set Control and Day0 as the baseline references
colData$Condition <- relevel(colData$Condition, ref = "Control")
colData$Day <- relevel(colData$Day, ref = "Day0")

# Combining everything into one dataframe
x <- DGEList(gene_expression, samples = colData)
```

Since we don't have raw count, we can not use voom. We will try to log transform it ourself

```{r}
exprs_data <- as.matrix(x$counts)
if(max(exprs_data, na.rm=TRUE) > 50) {
  exprs_data <- log2(exprs_data + 1)
}

```

Using Limma instead since we don't have raw data, and Deseq2 only take raw data

```{r}
library(limma)
library(Glimma) 
# unsupervised Clustering plot to see if the grouping on PCA space make sense. RUN THIS TO GET A FEEL OF THE DATA AND SEE IF ANY BATCH CORRECTIONS ARE NESSESARY
glMDSPlot(exprs_data, group = x$samples)

```

Creating Design matrix for both day and condition.

```{r}
Condition_Day <- x$samples$Condition_Day
Day <- x$samples$Day
design <- model.matrix(~Condition_Day)
colnames(design) <- gsub("Condition_Day", "", colnames(design))
colnames(design) <- gsub("DayDay", "Day", colnames(design))

view(design)
#The columns are summarized as the following:
# Intercept | ConditionApheresis | Day0 | Day2 | CondAph:Day0 | CondAph:Day2
# unbound intercept | 0 for control | * |  * | Combinations 
```

Contrast matrix

```{r}
contr.matrix <- makeContrasts(
  Apheresis_Day2vsDay1 = Apheresis_Day2 - Apheresis_Day1,
  Apheresis_Day1vsDay0 = Apheresis_Day1 - Apheresis_Day0,
  Apheresis_Day2vsDay0 = Apheresis_Day2 - Apheresis_Day0,
  Control_Day2vsDay1 = Control_Day2 - Control_Day1,
  Control_Day1vsDay0 = Control_Day1,
  Control_Day2vsDay0 = Control_Day2,
  Apheresis_Day0vsControl_Day0 = Apheresis_Day0,
  Apheresis_Day1vsControl_Day1 = Apheresis_Day1 - Control_Day1,
  Apheresis_Day2vsControl_Day2 = Apheresis_Day2 - Control_Day2,
  levels=colnames(design)
)
contr.matrix
```

```{r}
v <- voom(x, design, normalize.method = "none", plot=TRUE)
v
```

```{r}
vfit <- lmFit(x$counts, design)
vfit <- contrasts.fit(vfit, contrast = contr.matrix) 
efit <- eBayes(vfit, trend = TRUE) 
plotSA(efit)

```

```{r}
top_genes <- topTable(efit, coef = "Apheresis_Day0vsControl_Day0", adjust.method = "BH", number = Inf)
significant_top_genes <- top_genes[top_genes$adj.P.Val < 0.05 & top_genes$logFC > 1,]
print(significant_top_genes)
write_xlsx(cbind(Index = rownames(significant_top_genes), significant_top_genes), "~/Documents/School/Bio Stuff/Bios_20151/Bio_20251_Final_Project/Output_plot/Apheresis_Day0vsControl_Day0_Deg_Up.xlsx")
```

```{r}
# Summary of deg genes
up_genes <- efit$genes[decideTests(efit)[, "Apheresis_Day0vsControl_Day0"] == 1, ]
print(up_genes)
```

```{r}
volcanoplot(efit, coef = c(1), style = "p-value", highlight = 0, names = efit$genes$ID, hl.col="blue",
            xlab = "Log2 Fold Change", ylab = NULL, pch=16, cex=0.35)
abline(1,0, col = "red")
abline(0,-2, col= "blue")
abline(0,2, col= "blue")

```

```{r}
tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)
```

```{r}
glMDPlot(tfit, coef=1, status=de, main=colnames(efit)[1], side.main="ENTREZID", launch=TRUE)
```

TODO: Conversion to a single factor: Make Combined Design of condition and day (ie use Condition_Day).
